A Stopped Negative Binomial Distribution

### 2016-03-22 stat-and-prob-letters/snb-paper.tex
Yeah.

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{url}
\usepackage{subfig}
\DeclareMathOperator*{\argmin}{argmin}
\newcommand*{\argminl}{\argmin\limits}
\newtheorem{prop}{Proposition}

@@

Consider a prototypical Phase II single-arm clinical trial in which 12 patients
are enrolled and treated. If two or more respond then we reject the null
hypothesis that the treatment is not effective and the trial proceeds to the 
next stage. If fewer that two respond then the null is not rejected and the 
trial is terminated.

@@

Assume each of patient outcome can be modeled as an independent,
identically distributed Bernoulli($p$) random variable. The trial is realized
as a sequence of these random variables that stops when either a
specified number of success or failures is reached. In the
previous example suppose two successes were reached after enrolling 10
patients (one in the third step and one at the $10^{th}$). The sample
path is illustrated
in Fig.~\ref{fig:kane_viz}. The vertical axis denotes the number of
successful outcomes. The horizontal axis counts the number of patients that
have been enrolled. The horizontal and vertical boundaries represent
endpoints for the trial.

@@

To prove (\ref{eqn:pmf}), consider the
process $\mathbf{X} = \left\{X(k) : k = 0,1,... \right\}$
with $X(0)=0$ and
\begin{equation*} \label{eqn:proc}
X_{k+1} = X_k + b_{k+1} \ I_{\{ k-t < X_k < s\}}.
\end{equation*}
At each step a patient's outcome is measured. If it is success, the process 
advances one diagonally in the
positive horizontal and vertical direction.
Otherwise, it advances in the positive horizontal direction only. The
process continues until either $X_k = s$ or $X_k = k-t$.

@@

Next, define
\begin{equation} \label{stop_t}
S'(k, p, t) = {k-1 \choose k-t} p^{k-t} (1-p)^t
\end{equation}
and notice that $S(k, p, s) = S'(k, 1-p, s)$ by writing
${k-1 \choose k-s} = {k-1 \choose s-1}$.

@@

@@

\begin{figure}[p!]
\begin{center}
\includegraphics[width=\textwidth]{shapes.pdf}
\end{center}
\caption{Different shapes of the SNB distribution with parameters ($s$, $t$, $p$), as given. Red indicates mass contributed by hitting $s$, teal indicates
mass contributed by hitting $t$. \label{shapes.fig}}
\end{figure}

The SNB is a generalization of the negative 
binomial distribution. If $t$ is large then the $Y-s$ has a 
negative binomial distribution with
\begin{equation*}                                    %   (1)   
\mathbb{P}[Y=s+j]        \label{nb1.eq}          
  = {{s+j-1}\choose{s-1}} p^s (1-p)^j
\end{equation*}
for $\,j=0, 1,\ldots\,$. A similar statement can be made when $s$ is large
and $t$ is small. As a result, with proper parameter choice, the SNB
can mimic other probability distributions in a manner similar to 
those described in \cite{Peizer1968} and \cite{Best1974}. Examples are
shown in in Fig.~\ref{shapes.fig}.

For the special case of $\,s=t,$ the distribution of $\,Y\,$ is the
riff-shuffle, or minimum negative binomial distribution~\citep{Uppuluri1970}.
Similar derivations of the closely-related maximum negative binomial 
discrete distributions also appear in~\cite{Zhang2000}
and \cite{Zelterman2005}.
The maximum negative binomial is the smallest number of outcomes necessary to 
observe at least $s$ successes {\em and} $s$ failures. The SNB is the 
number of coin flips to observe {\em either} $s$ heads or $t$ tails.

@@

\begin{prop} \label{binomial_tail}
Let $Y$ be distributed as SNB($p$, $s$, $t$) and let 
$B$ be distributed Binomial with size $n=s+t-1$ and success probability
$p$. Then
\begin{equation}
\mathbb{P}[B \geq s] = \mathbb{P} [Y \leq n\ |\ \text{\#success} = s].
\end{equation}
That is, the probability that the number of successes is at least $s$
in the Binomial model is the same that the trial stops with $s$ 
successes in the SNB model.
\end{prop}
\begin{proof}
The Binomial tail probability is
\begin{align*}
\mathbb{P}[B \geq s] &= \sum_{k=s}^{s+t-1} {n \choose k} p^k (1-p)^{n-k} \\
  &= 1 - \sum_{k=0}^{s-1} {n \choose k} p^k (1-p)^{n-k} \\
  & 1 - \mathcal{I}_{1-p}.
\end{align*}
The corresponding SNB probability is
\begin{align*}
\mathbb{P} [Y \leq n\ |\ \text{\#success}=s] &= \sum_{k=s}^{s+t-1} S(k, p, s)\\
 &= {k-1 \choose s-1} p^s (1-p)^{k-s}.
\end{align*}
Let $i=k-s$. Use ${i+s-1 \choose s-1} = {i+s-1 \choose i}$
so the last summation can be rewritten as
\begin{align}
\mathbb{P} [Y \leq n\ |\ \text{\#success} = s] &= \sum_{i=0}^{t-1} 
  {i+s-1 \choose i} p^s (1-p)^i\\
  &= 1 - \mathcal{I}_{1-p}(t, s)
\end{align}
completing the proof.
\end{proof}

Fig.~\ref{fig:snb_bin_compare} shows the case
where $s=2$, $t=12$, and $p=0.2$. The probability masses represented in
red are equal as are the masses in teal. The probability that $s$
successes are reached in the SNB process is the same as the binomial 
probability of at least two successes. Likewise, the probability that $t$ 
failures are reached in the SNB process is the same as the binomial
probability of zero or one successes.

\begin{figure}[t!]
\centering
\subfloat[The SNB distribution.]{\includegraphics[width=0.5\textwidth]{snb_density.pdf}}
\hfill
\subfloat[The Binomial distribution.]{\includegraphics[width=0.5\textwidth]{bin_density.pdf}}
\caption{
SNB(0.2, 2, 12) with mass contributed from 
$s$ successes (red) or $t$ failures (teal) along with Bin(0.2, 13) with
at least 2 successes (red) or fewer (teal).
}
\label{fig:snb_bin_compare}
\end{figure}

@@

@@

Let us consider a Bayesian analysis where there is a Beta prior distribution
on $p$.
\begin{prop}
The posterior PMF of the Stopped Negative Binomial distribution with a Beta($\alpha$, $\beta$) prior is:
\begin{align} \label{eqn:posterior}
f(k | s, t, \alpha, \beta) &= 
  {k-1 \choose s-1} \frac{B\left(\alpha+s, k-s+\beta \right)}{B(\alpha, \beta)} 
    \ I_{\{s \leq k \leq s+k-1\}} \ + \nonumber \\
  & {k-1 \choose k-t} 
    \frac{B\left(\alpha + k - t, t+\beta\right)}{B(\alpha, \beta)} 
    \ I_{\{t \leq k \leq s+k-1\}}.
\end{align}
\end{prop}
\begin{proof}
For notational simplicity, assume that $s,t \leq k \leq s+t-1$. When this is not the case appropriate terms should be removed as dictated by the indicator functions.
\begin{align*}
f(k | s, t, \alpha, \beta) = \frac{1}{B(\alpha, \beta)} & \int_0^1 {k-1 \choose s-1} p^{\alpha +s -1} \left(1-p\right)^{k-s+\beta-1} + \\
 & {k-1 \choose k-t} p^{k-t+\alpha-1}\left(1-p\right)^{t+b-1} dp \\
= \frac{1}{B(\alpha, \beta)}  {k-1 \choose s-1} & \int_0^1  p^{\alpha +s -1} \left(1-p\right)^{k-s+\beta-1} dp + \\
 & \frac{1}{B(\alpha, \beta)} {k-1 \choose k-t} \int_0^1  p^{k-t+\alpha-1}\left(1-p\right)^{t+b-1} dp
\end{align*}
The result follows by applying the definition of the Beta function to the integral terms.
\end{proof}

\section{Discussion and Conclusion}

We have presented a new discrete distribution by curtailed sampling rules 
common in early-stage clinical trials, which we refer to as the Stopped 
Negative Binomial distribution. The distribution models the stopping time of 
a sequential trial where the trial is stopped when a number of events are 
accumulated. A connection between the Binomial tail probability and the SNB
distribution was shown; It's MGF was derived; and
the posterior distribution was derived for the case when the 
event probability $p$ has a Beta distribution.

Current work focuses on two different areas. First, the distribution
is being applied to clinical trial design
where the number of enrollees is smaller than traditional trials.
This scenario is increasingly common in targeted cancer
therapy where treatment may only be appropriate for small subpopulations.
Second, generalizations and extensions of the distribution are being pursued
for applications power analysis. The posterior construction allows us
to quantify the uncertainty both in the response probability as well as the
time until a trial is complete. This suggests new approaches to sample
size calculations as well as trial monitoring.

>>>>

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{url}
\usepackage{subfig}
\DeclareMathOperator*{\argmin}{argmin}
\newcommand*{\argminl}{\argmin\limits}
\newtheorem{prop}{Proposition}
\newtheorem{corollary}{Corollary}

@@

Consider a prototypical Phase II single-arm clinical trial in which 12 patients
are enrolled and treated. If two or more respond then we reject the null
hypothesis that the treatment is not effective and the trial is deemed
successful. If fewer that two respond then the null is not rejected and the 
trial is terminated.

@@

Assume each patient outcome can be modeled as an independent,
identically distributed Bernoulli($p$) random variable. The trial is realized
as a sequence of these random variables that stops when either a
specified number of success or failures is reached. In the
previous example suppose two successes were reached after enrolling 10
patients (one in the third step and one at the $10^{th}$). The sample
path is illustrated
in Fig.~\ref{fig:kane_viz}. The vertical axis denotes the number of
successful outcomes. The horizontal axis counts the number of patients that
have been enrolled. The horizontal and vertical boundaries represent
endpoints for the trial.

@@

To prove (\ref{eqn:pmf}), consider the
process $\mathbf{X} = \left\{X(k) : k = 0,1,... \right\}$
with $X(0)=0$ and
\begin{equation*} \label{eqn:proc}
X_{k+1} = X_k + b_{k+1} \ I_{\{ k-t < X_k < s\}}.
\end{equation*}
At each step a patient's outcome is measured. If it is a success, the process 
advances one diagonally in the
positive horizontal and vertical direction.
Otherwise, it advances in the positive horizontal direction only. The
process continues until either $X_k = s$ or $X_k = k-t$.

@@

Next, define
\begin{equation} \label{stop_t}
S'(k, p, t) = {k-1 \choose k-t} p^{k-t} (1-p)^t
\end{equation}
and notice that $S(k, p, s) = S'(k, 1-p, s)$ by writing
\begin{equation*}
{k-1 \choose k-s} = {k-1 \choose s-1}.
\end{equation*}

@@

\begin{corollary} \label{conditional_distribution}
Assume a trial that stops on the $s$th successs or $t$failure as before. 
Furthermore, assume we observer $s'$ successes and $t'$ failures
where $s' < s$ and $t' < t$.
That is, we see the number of success and failures at a time before
reaching an endpoint. Then the number of subsequent enrollements needed 
to reach an endpoint is SNB($p$, $s-s'$, $t-t'$).
\end{corollary}
\begin{proof}
The result follows directly by realizing that, conditioned on $s'$
success and $t'$ failures, there are $s-s'$ successes needed to reach the
success endpoint and $t-t'$ failures needed to reach the failure endpoint.
\end{proof}

@@

\begin{figure}[p!]
\begin{center}
\includegraphics[width=\textwidth]{shapes.pdf}
\end{center}
\caption{Different shapes of the SNB distribution with parameters ($s$, $t$, $p$), as given. Black indicates mass contributed by hitting $s$, grey indicates
mass contributed by hitting $t$. \label{shapes.fig}}
\end{figure}

The SNB is a generalization of the negative 
binomial distribution. If $t$ is large then $Y-s$ has a 
negative binomial distribution with
\begin{equation*}                                    %   (1)   
\mathbb{P}[Y=s+j]        \label{nb1.eq}          
  = {{s+j-1}\choose{s-1}} p^s (1-p)^j
\end{equation*}
for $\,j=0, 1,\ldots\,$. A similar statement can be made when $s$ is large
and $t$ is small. As a result, with proper parameter choice, the SNB
can mimic other probability distributions in a manner similar to 
those described in \cite{Peizer1968} and \cite{Best1974}. Examples are
shown in in Fig.~\ref{shapes.fig}.

For the special case of $\,s=t,$ the distribution of $\,Y\,$ is the
riff-shuffle, or minimum negative binomial distribution~\citep{Uppuluri1970}.
Similar derivations of the closely-related maximum negative binomial 
discrete distributions also appear in~\cite{Zhang2000}
and \cite{Zelterman2005}.
The maximum negative binomial is the smallest number of outcomes necessary to 
observe at least $s$ successes {\em and} $s$ failures. The SNB is the 
number of outcomes to observe {\em either} $s$ successes or $t$ failures.

@@

The probability of reaching the success endpoint in an 
SNB($p$, $s$, $t$) random variable is 
equal to the probability of at least $s$ successes in a Binomial distributed 
random variable with size $s+t-1$ and success probability $p$. 
Likewise, the probability of reaching the failure endpoint is equal
to at most $s-1$ successes in a Binomial distribution.
\begin{prop} \label{binomial_tail}
Let $Y$ be distributed as SNB($p$, $s$, $t$) and let 
$B$ be distributed Binomial with size $n=s+t-1$ and success probability
$p$. Then
\begin{equation}
\mathbb{P}[B \geq s] = \mathbb{P} [Y \leq n\ |\ \text{\#success} = s].
\end{equation}
That is, the probability that the number of successes is at least $s$
in the Binomial model is the same that the trial stops with $s$ 
successes in the SNB model.
\end{prop}
\begin{proof}
The Binomial tail probability is
\begin{align*}
\mathbb{P}[B \geq s] &= \sum_{k=s}^{s+t-1} {n \choose k} p^k (1-p)^{n-k} \\
  &= 1 - \sum_{k=0}^{s-1} {n \choose k} p^k (1-p)^{n-k} \\
  & 1 - \mathcal{I}_{1-p}.
\end{align*}
The corresponding SNB probability is
\begin{align*}
\mathbb{P} [Y \leq n\ |\ \text{\#success}=s] &= \sum_{k=s}^{s+t-1} S(k, p, s)\\
 &= {k-1 \choose s-1} p^s (1-p)^{k-s}.
\end{align*}
Let $i=k-s$. Use ${i+s-1 \choose s-1} = {i+s-1 \choose i}$
so the last summation can be rewritten as
\begin{align}
\mathbb{P} [Y \leq n\ |\ \text{\#success} = s] &= \sum_{i=0}^{t-1} 
  {i+s-1 \choose i} p^s (1-p)^i\\
  &= 1 - \mathcal{I}_{1-p}(t, s)
\end{align}
completing the proof.
\end{proof}

Fig.~\ref{fig:snb_bin_compare} shows the case
where $s=2$, $t=12$, and $p=0.2$. The probability masses represented in
black are equal as are the masses in grey. The probability that $s$
successes are reached in the SNB process is the same as the binomial 
probability of at least two successes. Likewise, the probability that $t$ 
failures are reached in the SNB process is the same as the binomial
probability of zero or one successes.

\begin{figure}[t!]
\centering
\subfloat[The SNB distribution.]{\includegraphics[width=0.5\textwidth]{snb_density.pdf}}
\hfill
\subfloat[The Binomial distribution.]{\includegraphics[width=0.5\textwidth]{bin_density.pdf}}
\caption{
SNB(0.2, 2, 12) with mass contributed from 
$s$ successes (black) or $t$ failures (grey) along with Bin(0.2, 13) with
at least 2 successes (black) or fewer (grey).
}
\label{fig:snb_bin_compare}
\end{figure}

@@

The moment generating function is calculated in a manner similar to 
that of two negative binomial distributions. However, in the negative
binomial derivation the
term involving the regularized incomplete beta function integrates to
one since it is integrated over it's entire support. This is not the
case in the SNB and it appears in both terms.

@@

Let us consider a Bayesian analysis where there is a Beta prior distribution
on $p$.
\begin{prop} \label{prop:bayesian}
The posterior PMF of the Stopped Negative Binomial distribution with a Beta($\alpha$, $\beta$) prior is:
\begin{align} \label{eqn:posterior}
f(k | s, t, \alpha, \beta) &= 
  {k-1 \choose s-1} \frac{B\left(\alpha+s, k-s+\beta \right)}{B(\alpha, \beta)} 
    \ I_{\{s \leq k \leq s+k-1\}} \ + \nonumber \\
  & {k-1 \choose k-t} 
    \frac{B\left(\alpha + k - t, t+\beta\right)}{B(\alpha, \beta)} 
    \ I_{\{t \leq k \leq s+k-1\}}.
\end{align}
\end{prop}
\begin{proof}
For notational simplicity, assume that $s,t \leq k \leq s+t-1$. When this is not the case appropriate terms should be removed as dictated by the indicator functions.
\begin{align*}
f(k | s, t, \alpha, \beta) = \frac{1}{B(\alpha, \beta)} & \int_0^1 {k-1 \choose s-1} p^{\alpha +s -1} \left(1-p\right)^{k-s+\beta-1} + \\
 & {k-1 \choose k-t} p^{k-t+\alpha-1}\left(1-p\right)^{t+b-1} dp \\
= \frac{1}{B(\alpha, \beta)}  {k-1 \choose s-1} & \int_0^1  p^{\alpha +s -1} \left(1-p\right)^{k-s+\beta-1} dp + \\
 & \frac{1}{B(\alpha, \beta)} {k-1 \choose k-t} \int_0^1  p^{k-t+\alpha-1}\left(1-p\right)^{t+b-1} dp
\end{align*}
The result follows by applying the definition of the Beta function to the integral terms.
\end{proof}

\section{Discussion}

The SNB can be used to extend clinical trial methodology
in at least two areas. First, it can be used to construct nested criteria
for early stopping and efficacy. These trials consist of two-stages.
If the failure endpoint is reached in the first stage then the 
trial stops due to futility. Otherwise, the trial enters the second 
stage, which consists of responders from the first stage along with
new enrollees in the second. Thus the design provides
both this provide early testing for futility along with the ability
to make statistically sound decisions sooner than traditional trials.

A second avenue for methodological innovation comes directly from
Corollary \ref{conditional_distribution} and Proposition \ref{prop:bayesian}.
While a trial is being conducted the response probabilities can be fitted
and updated as new results are received. The number of individual successes
and failure are used to learn the shape characteristics of the corresponding
Beta distribution and resulting posterior
distribution. An estimate of the probability of reaching the success (and
failure) endpoints can be provided at each step of the trial.

There are also several avenues for extending the SNB itself. First is the 
development of a stopped negative multinomial distribution allowing
for more than two endpoints. This extension would have clinical trial
applications including dose escalation where the outcomes are response
to a therapy, failure to respond, and an adverse outcome endpoint where
and enrollee is removed from the trial with neither response or non-response
due to side effects.

A second, more theoretical extension is to the asymptotic setting. Under
proper reparameterization, the process from which the SNB is derived 
can be shown to converge to a Brownian motion. In this case, we conjecture
that the SNB can be defined in terms of the L\'{e}vy-Bachelier 
\citep{Bachelier1900} formula
thereby relating the SNB to the classic result in stochastic calculus.
